---
title: "Creating Azure Logic Apps from R"
author: "Christoph"
date: 2019-05-19T21:13:14-05:00
categories: ["R", "Azure"]
tags: ["Logic Apps", "serverless"]
---

Logic Apps is a serverless framework in Azure quite similar to [IFTTT](ifttt.com) (if this, then that) and [Zapier](zapier.com) that allows you to connect different services and create workflows. 

You can define different types of triggers based on:

- time and
- events (e.g. http requests, messages received, ...)

that start workflows.

A super simple example is, you receive an email and this event triggers an automatic reply.

Logic Apps can be created using a visual interface in the Azure Portal, but also programmatically using the Azure REST api. In this post I want to show you how we can create Logic Apps in Azure directly from R. 

## Getting Started
To create or update Logic Apps, we will use the following REST call template:
```
PUT subscriptions/<subscriptionId>/resourceGroups/<resourceGroupName>/providers/Microsoft.Logic/workflows/<workflowName>?api-version=2016-06-01  HTTP/1.1
Authorization: Bearer <bearer-token>
Content-Type: application/json
Host: management.azure.com

{
  <body>
}
```

We will use `httr` to send the PUT request, but first we need to acquire an access token:
```{r}
library(httr)

app_name = "logic_app"
# client_id of Azure CLI
client_id = "04b07795-8ddb-461a-bbee-02f9e1bf7b46"
client_secret = NULL
resource_uri = "https://management.core.windows.net/"

# Endpoints are of the form: 
# https://login.windows.net/<common | tenant-id>/oauth2/<authorize | token>
azure_endpoint = oauth_endpoint(authorize = "https://login.windows.net/441ec6d2-e277-43f8-b599-8dc54649d5c8/oauth2/authorize",
                                access = "https://login.windows.net/441ec6d2-e277-43f8-b599-8dc54649d5c8/oauth2/token")

logic_app = oauth_app(
  appname = app_name,
  key = client_id,
  secret = client_secret
)

token <- oauth2.0_token(azure_endpoint, logic_app,
  user_params = list(resource = resource_uri),
  use_oob = FALSE
)
```

Now, let's try to stitch together our REST call to create a simple Logic App:
```{r}
url = "https://management.azure.com/subscriptions/a946be7b-3533-413d-9366-91e10c0b209b/resourceGroups/serverless/providers/Microsoft.Logic/workflows/serverless_testworkflow?api-version=2016-06-01"

body = jsonlite::fromJSON(txt = "../../data/logic_app.json")

PUT(url = url, 
    body = body,
    add_headers(Authorization = paste0("Bearer ", token$credentials$access_token))
    )
```



https://docs.microsoft.com/de-at/rest/api/logic/workflows/createorupdate

https://docs.microsoft.com/en-us/rest/api/azure/

PUT https://management.azure.com/subscriptions/{subscriptionId}/resourceGroups/{resourceGroupName}/providers/Microsoft.Logic/workflows/{workflowName}?api-version=2016-06-01
Authorization: 
Content-Type: application/json


## Open
- Common geht nicht bei login.windows.net nur tenant-id?