---
title: "Creating Azure Logic Apps from R using httr"
author: "Christoph"
date: 2019-05-30T14:41:14-05:00
categories: ["R", "Azure"]
tags: ["Logic Apps", "serverless"]
---

Logic Apps is a serverless framework in Azure quite similar to [IFTTT](ifttt.com) (if this, then that) and [Zapier](zapier.com) that allows you to connect different services and create workflows. 

You can define different types of triggers based on:

- time and
- events (e.g. http requests, messages received, ...)

to start workflows.

Logic Apps can be created using a visual interface in the Azure Portal, but also programmatically using the Azure REST api. In this post I want to show you how we can create Logic Apps in Azure directly from R.

## Getting Started
To create or update Logic Apps, we will use the following REST call template:
```
# Line breaks for legibility only
PUT subscriptions/<subscriptionId>/resourceGroups/<resourceGroupName>/
providers/Microsoft.Logic/workflows/<workflowName>
?api-version=2016-06-01  HTTP/1.1
Authorization: Bearer <bearer-token>
Content-Type: application/json
Host: management.azure.com

{
  <body>
}
```

But before we can start, we need to create a resource group in Azure for our Logic App.

We will use `httr` to send the PUT request to create the resource group, but first we need to acquire an access token (Note: make sure you have `httpuv` installed as well):
```{r, eval = FALSE}
library(httr)

app_name = "azure_mgmt"

# client_id of Azure CLI so we do not need to register an app
client_id = "04b07795-8ddb-461a-bbee-02f9e1bf7b46"    

# we have no client secret, we do everything interactively
client_secret = NULL

# Your tenant_id/directory_id -> azure portal/azure active directory/Properties/directory id
tenant_id = "441ec6d2-e277-43f8-b599-8dc54649d5c8"

# azure subscription -> azure portal/subscriptions
subscription = "a946be7b-3533-413d-9366-91e10c0b209b" 

# name of the logic app workflow in azure
workflow_name = "serverless_testworkflow"     

# endpoint we need to request permission to
resource_uri = "https://management.core.windows.net/"

# Endpoints are of the form: 
# https://login.windows.net/<tenant-id>/oauth2/<authorize | token>
azure_endpoint = oauth_endpoint(authorize = glue::glue("https://login.windows.net/{tenant_id}/oauth2/authorize"),
                                access = glue::glue("https://login.windows.net/{tenant_id}/oauth2/token"))

logic_app = oauth_app(
  appname = app_name,
  key = client_id,
  secret = client_secret
)

token <- oauth2.0_token(azure_endpoint, logic_app,
  user_params = list(resource = resource_uri),
  use_oob = FALSE
)
```

We can now create our resource group like so:
```{r, eval = FALSE}
resource_group = "serverless"                         # Resource group we are going to use
url = glue::glue("https://management.azure.com/subscriptions/{subscription}/resourcegroups/{resource_group}?api-version=2018-05-01")

body = paste0('
  {
  "location": "westeurope"
}'
)

PUT(url = url,
    add_headers(.headers = c(
                "Authorization" = paste0("Bearer ", token$credentials$access_token)
                ,"Content-Type" = "application/json")
    ),
    body = body,
    encode = "json"
    )
```

Now, let's try to stitch together our REST call to create a simple Logic App.

We will setup a super simple logic app with an http trigger that returns a "Hello world!":

```
{
	"properties": {
	    "definition": {
	        "$schema": "https://schema.management.azure.com/providers/Microsoft.Logic/schemas/2016-06-01/workflowdefinition.json#",
	        "actions": {
	            "Response": {
	                "inputs": {
	                    "body": {
	                        "response": "Hello world!"
	                    },
	                    "statusCode": 200
	                },
	                "kind": "Http",
	                "runAfter": {},
	                "type": "Response"
	            }
	        },
	        "contentVersion": "1.0.0.0",
	        "outputs": {},
	        "parameters": {},
	        "triggers": {
	            "manual": {
	                "inputs": {
	                    "schema": {}
	                },
	                "kind": "Http",
	                "type": "Request"
	            }
	        }
	    }
	},
	    "location": "westeurope"
}

```

Let's build the query to submit the request:
```{r, eval = FALSE}
url = glue::glue("https://management.azure.com/subscriptions/{subscription}/resourceGroups/{resource_group}/providers/Microsoft.Logic/workflows/{workflow_name}?api-version=2016-06-01")

body = readLines(con = "../../data/logic_app.json")


PUT(url = url, 
    add_headers(.headers = c(
                "Authorization" = paste0("Bearer ", token$credentials$access_token)
                ,"Content-Type" = "application/json")
    ),
    body = body,
    encode = "json"
    )
```
And we are done:) If you go to your resource group in the Azure portal, you should see the logic app we defined above.


## Links and more information
You can find more detailed information in the official Microsoft Azure docs:

- [Logic Apps - Official Docs](https://docs.microsoft.com/de-at/rest/api/logic/workflows/createorupdate)
- [Azure REST api](https://docs.microsoft.com/en-us/rest/api/azure/)