---
title: "Using data.table deep copy"
author: "Christoph"
date: 2018-05-12T14:41:14-05:00
categories: ["R"]
tags: ["data.table", "R", "change by reference"]
---

`data.table` is an awesome R package, but there are a few things you need to watch out for when using it.

R usually does not modify objects in place (e.g. by reference), but makes a copy when you change a value and saves this copy. This can be a problem if you work with large datasets since you will quickly need about 2-3x your dataset size
in RAM to accommodate these intermediary copies.

With `data.table` you can modify objects by reference which is:

- way faster (you do not need to allocate new memory and copy everything) and
- uses way less RAM.

Now consider the following example:
```{r}
library(data.table)

dt_mtcars = mtcars

## change to data.table object
setDT(dt_mtcars)

## assign to new variable
dt_mtcars_shallow_copy = dt_mtcars
```

What do you think will happen when we modify `dt_mtcars_shallow_copy` by reference?
```{r}
# Let's drop a column in dt_mtcars_shallow_copy
dt_mtcars_shallow_copy[,cyl := NULL]

# Are the two objects identical?
identical(dt_mtcars_shallow_copy, dt_mtcars)
```


They are! How could that happen? Let's check the memory location in RAM where our objects are stored:
```{r}
tracemem(dt_mtcars)
tracemem(dt_mtcars_shallow_copy)
```


Both objects are stored at the same place in our computer's memory. The reason for this is actually quite simple:
When you do an assignment in R, both variables point to the same location in memory until one of them changes. When we change one
of the variables using R's normal functions, R makes a copy of the object. However, since we used `data.table` to modify the
object by reference R never makes a copy and both variables till point to the same address with the changed dataset. 

You can force a deep copy (aka copy an object to its own location in memory) by using `copy()`:
```{r}
# use copy() to create a deep copy
dt_mtcars_deep_copy = copy(dt_mtcars)

tracemem(dt_mtcars_deep_copy)
tracemem(dt_mtcars)
```

That's it, hope you found the post useful:)