---
title: "An Introduction to Statistical Learning"
author: "Christoph"
date: 2020-02-09T14:41:14-05:00
categories: ["R", "Theory"]
tags: ["R"]
---

In this blog post I want to give a brief overive of the core ideas behind the 
statistical learning framework and show you how to implement a linear regression 
model based on this framework.

This post is based on the excellent work "Tree Boosting With XGBoost - Why Does 
XGBoost Win 'Every' machine Learning Competition" by Didrik Nielson.

First, let's start with some general background info: Statistical learning refers 
to "learning" patterns from data that generalize well to new incoming data. We can 
distinguish between two main areas, supervised and unsupervised learning, but in 
this post I will only focus on the former.

In a supervised learning problem, we have a variable we want to predict ($y$) and some 
covariates, also called features ($X$) that might help explain changes in $y$. 
Mathematically speaking we have:

$$
\overbrace{\mathcal{D}}^{data set} = \{(y_1, X_1), (y_2, X_2), ..., (y_n, Xn)\} \\
\text{with} \\
y \in \mathcal{Y} \text{ and } X = (X_1, ... X_p) \in \mathcal{X} 
$$
We want to find a function $f$ that helps us predict $Y$ given $X$ with high 'accuracy'. 
the function that defines how we measure accuracy is called **loss function** and our 
goal is to minimize the **expected loss**, also know as the **risk**. The function 
minimizing risk is called **target function** and is the function we want to find. 

So we have:
$$
L: \mathcal{Y} \times \mathcal{A} \to \mathbb{R}_+ \text{... loss function}\\
\text{with}\\
a \in \mathcal{A} \text{... action from set of allowable actions, i.e. a prediction } \hat{y}
$$

Some examples include:

- Squared loss: $L(y, a) = \frac{1}{2}(y - a)^2$ 
- Absolute loss: $L(y, a) = |(y - a)|$

The risk is now simply the expected value of the loss function:
$$
R(a) = \mathbb{E}[L(Y,a)]
$$
The optimal action $a^*$ is given by:
$$
a^* = \underset{a \in \mathcal{A}}{\text{argmin }} R(a)
$$

> Note: The argmin operator returns the point where the minimum value is reached, while 
the min operator returns the value of the function at that point!

To generate predictions $a$, we use a function, i.e. a model, that maps $X$ to a:
$$
f: \mathcal{X} \to \mathcal{A} \\
a = f(x)
$$
Where $\mathcal{A}^{\mathcal{X}}$ is the set of all functions mapping $\mathcal{X}$ 
to $\mathcal{A}$.

The risk of a model is therefore simply:
$$
R(f) = \mathbb{E}[L(Y,f(X))] = \mathbb{E}[\mathbb{E}[L(Y, f(X)) | X]]
$$
by the law of total expectation, with the optimal model given by:

$$
f^* = \underset{f \in \mathcal{A}^{\mathcal{X}}}{\text{argmin }} R(f) \text{ ... target function}
$$

This is the function we want to find/estimate. We can calculate the target function pointwise:

$$
f^*(X) =\text{argmin }_{f(x) \in \mathcal{A}^{\mathcal{X}}} \mathbb{E}[L(Y,f(X)) | X = x] \text{ } \forall x \in \mathcal{X}
$$
Let's look at the squared loss again. Squared loss risk is given by:
$$
R(f) = \mathbb{E}[(Y - f(x))^2 | X = x]
$$
We can show that the target function is the conditional expectation:
$$
f^*(x) = \mathbb{E}[Y|X = x]
$$
and the risk of the target function is given by the conditional variance:
$$
R(f^*(x)) = Var[Y | X = x]
$$

Since we don't know the true model risk, we use the empirical risk as a proxy, which is 
simply an estimate of the true risk:
$$
\hat{R(f)} = \frac{1}{n}\sum_{i=1}^nL(y_i, f(x_i)) \\
\text{with} \\
lim_{n \to \infty}\hat{R(f)} = R(f) \text{ by the law of large numbers}
$$

So the empirical risk minimizer is given by:
$$
\hat{f} = argmin_{f \in \mathcal{F} \subseteq \mathcal{A}^{\mathcal{X}}}\hat{R(f)}
$$
> Note: We consider only a subspace $\mathcal{F} \subseteq \mathcal{A}^{\mathcal{X}}$

Since we only have finitely many datapoints, but an infinite function search space, there are 
infinitely many functions taht fit perfectly in-sample. We thus need to impose additional constraints 
on our functions, which leads us to consider different model classes:

A model class is nothing more than a restriction $\mathcal{F} \subseteq \mathcal{A}^{\mathcal{X}}$ of the 
total function space and is sometimes also called hypothesis space. Each $mathcal{F}$ can be 
viewed as a model class. Take for example the class of linear models:

$$
\mathcal{F} = \{f: f(x) = \theta_0 + \sum_{i = 1}^p \theta_ixi, \forall x \in \mathcal{X}}\}
$$

To sum up: empirical risk minimization & a given model class turns "learning" into function optimization.

Now, let's implement linear regression in the statistical learning framework. We want to predict miles per gallon (mpg) using features from the mtcars dataset:
```{r}
library(ggplot2)
mtcars
```

Let's look at the relation between horsepower (hp) and miles per gallon (mpg):
```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship between horsepower and mpg",
       subtitle = "More horsepower seems to imply less mpg")
```


```{r}
empirical_risk = function(loss_fun, y, yhat) {
  mean(loss_fun(y, yhat))
}

squared_loss = function(y, yhat) (y - yhat)^2

optim_wrapper = function(params) {
  
  empirical_risk(squared_loss, y = mtcars$mpg, yhat = params[1] + params[2]*mtcars$hp)
  
}

par_loss_sqrt = optim(par = c(0,0), fn = optim_wrapper)
par_loss_sqrt
```

Let's compare our output to R's `lm` function:

```{r}
lm(mpg ~ hp, data = mtcars)
```
We get almost the same parameters, which was to be expected, because linear regression or ordinary least squares (OLS) relies on squared loss as the name implies:)

Now, let's run the same experiment, with absolute error as a loss function:
```{r}
absolute_loss = function(y, yhat) abs(y - yhat)

optim_wrapper = function(params) {
  
  empirical_risk(absolute_loss, y = mtcars$mpg, yhat = params[1] + params[2]*mtcars$hp)
  
}

par_loss_abs = optim(par = c(0,0), fn = optim_wrapper)
par_loss_abs
```

And as theory predicts, parameters change, because our risk minimizer is now the conditional median.

We can also train a non-linear model:
```{r}
optim_wrapper = function(params) {
  
  empirical_risk(squared_loss, y = mtcars$mpg, yhat = params[1]*exp(params[2]*mtcars$hp))
  
}

par_custom = optim(par = c(0, 0), fn = optim_wrapper, method = "SANN", control = list(maxit = 800000))
par_custom
```

Even though this is quite a simple model, just using the default optimization settings as I did before does not cut it anymore. The default algorithms are extremely sensitive to the starting values you pick, but they do work if you take a look at the graph and provide 'good' initial guesses. Moreover, since fitting an exponential decay function relys heavily on getting the boundaries correct, a squared loss might be more appropriate than absolute loss as extreme errors are penalized more strongly.

Let's take a look at the model output:
```{r}
df_plot = data.frame(mpg = mtcars$mpg, 
                     hp = mtcars$hp,
                     yhat_sqrt = par_loss_sqrt$par[1] + par_loss_sqrt$par[2] * mtcars$hp,
                     yhat_abs = par_loss_abs$par[1] + par_loss_abs$par[2] * mtcars$hp,
                     yhat_cust = par_custom$par[1]*exp(par_custom$par[2] * mtcars$hp)
                    )

ggplot(df_plot, aes(x  = hp, y = mpg)) +
  geom_point() +
  geom_line(aes(y = yhat_sqrt)) + 
  geom_line(aes(y = yhat_abs)) +
  geom_line(aes(y = yhat_cust), color = "red") +
  labs(title = "Results using different optimization algorithms",
       subtitle = "Red: exponential decay model with squared loss, black: linear models")
```


While the examples above are very simple, the same logics applies to arbitrarily complex problems. In practice, the main problem is finding a good optimizer to solve the optimization problem as we already saw with the last example.

Sidenote: Terminology in this field can be quite confusing. I was spending quite some time trying to figure out, if I should be talking about objective, loss or cost functions. You can take a look at this [stats.stackexchange](https://stats.stackexchange.com/questions/179026/objective-function-cost-function-loss-function-are-they-the-same-thing) discussion  to see different ways people use these terms in practice. Personally, I like the definition that was accepted as an answer, which goes like this:

- *Loss function*: defined on a data point and measures the penalty (e.g. squared loss)
- *Cost function*: usually more general, might be the sum of loss functions over a training set plus some penalty (e.g. MSE)
- *Objective function*: the most general term for any function that you want to optimize

User lejlot summarises with:
> A loss function is a part of a cost function which is a type of an objective function.

But the other definitions are quite nice as well:)

I hope you found my post helpful!

# TODO
- Check notation with dataset
- Add likelihood based loss -> normal: https://medium.com/quick-code/maximum-likelihood-estimation-for-regression-65f9c99f815d
https://stats.stackexchange.com/questions/252267/the-loss-function-of-the-gaussian-maximum-likelihood-estimator
- Check entire notation
- Add relationship: loss <> risk: conditional variance, MSE