---
title: "Linear Algebra for Statistics"
author: "Christoph"
date: 2020-04-11T14:41:14-05:00
categories: ["theory"]
tags: ["math", "statistics"]
---

Since quite some time has passed since I took my linear algebra courses, I thought I could comb through my old course notes and write a small post about linear algebra stuff that is quite useful to remember for statistics.

```{r, include=FALSE}
library(ggplot2)
```

## Some useful Properties of Matrices

### Some Calculation Rules
Let $A$, $B$, $C$ be $(m \times n)$ matrices and let $\alpha, \beta \in \mathbb{R}$. Then:

1. $(A + B)^t = A^t + B^t$
2. $(AB)^t = B^t A^t$
3. $A(BC) = (AB)C$
4. $A(B + C) = AB + AC$
5. $(A + B)C = AC + BC$
6. $AB \neq BA \text{ in general}$
7. If $A, B$ invertible, then: 
  1. $(A^{-1})^{-1} = 1$
  2. $(A^t)^{-1} = (A^{-1})^t$
  3. $(AB)^{-1} = B^{-1}A^{-1}$

### Some Useful Properties

- Symmetric: A matrix $A$ is called symmetric $\iff$ $A = A^t$
- Inverse: Matrices $A, B$ are inverse to each other $iff$ $AB = BA = I$.
  $A, B$ are called regular or invertible: $B = A^{-1}$. $A$ is called singular if it is not invertible
- Let $A$ be a square matrix. The following are equivalent:
  1. $A$ is invertible
  2. $Ax = 0 \iff x = 0$
  3. The rows/columns of $A$ are linearly independent

## Transformation Matrix

## Change of Basis
When we work with vectors, we generally use the [Cartesian Coordinates](https://en.wikipedia.org/wiki/Cartesian_coordinate_system#Representing_a_vector_in_the_standard_basis), i.e. we represent vectors wrt. the standard basis $E$. Suppose we have a vector $v = (1, 2)$. Then $x = 1$ and $y = 2$ are the coordinates of $v$ wrt. the standard basis:

$$
E = \begin{bmatrix}
            1 &0 \\
            0 &1
      \end{bmatrix} \\
v = x \begin{bmatrix}
            1 \\
            0
      \end{bmatrix} + 
    y \begin{bmatrix}
            0 \\
            1
      \end{bmatrix} = v^t E = Ev
$$

```{r}
plot(x = 1, y = 2, xlab = "x", ylab = "y")
```
- http://www.boris-belousov.net/2016/05/31/change-of-basis/
- relation between change of basis and projection onto a basis
https://math.stackexchange.com/questions/2963797/relation-between-change-of-basis-and-projection-onto-a-basis

Special case: change of basis in same vector space! endomorphism (covariance?)


## Matrices as Linear Transformations
### Scaling
### Rotation
### Shearing
### Reflection
### Orthogonal Projection

## Some Important Matrices
### Covariance Matrix


### Jacobian Matrix -> not really linalg related?
- Matrix Calculus https://en.wikipedia.org/wiki/Matrix_calculus#Derivatives_with_vectors

## References
- http://ibgwww.colorado.edu/~carey/p7291dir/handouts/matrix.algebra.pdf