---
title: "Working with REST APIs for Data Scientists in R"
author: "Christoph"
date: 2019-04-08T14:41:14-05:00
categories: ["R", "Python", "API"]
tags: ["REST", "OAuth"]
---


With the growing importance of cloud computing more and more services are 
exposed as REST apis. In this post, I want to give a hands-on introduction for data 
scientists from non-software-engineering backgrounds on how to work with REST apis.

But before we dive straight into the code, let's start with some background information:

## A (short) history of URLs and REST
REST is an abbreviation for 'REpresentational State Transfer' and '[defines a set of 
constraints to be used for creating Web services](https://en.wikipedia.org/wiki/Representational_state_transfer)'. 
A web service that implements all REST architecture patterns is called RESTful. With REST apis we can 
access and change web resources. Note that while most REST apis use 'http' as their protocol, in theory you could 
use a different protocol as well. A web resource can be anything that has a uniform resource 
identifier also called URI. A uniform resource locator, also called URL or simple web address, 
is probably the most common kind of URI. A URL tells us where a particular resource is located by telling us its 
primary access mechanism (e.g. the 'http' protocol).
A URI is simply an identifier of a resource that in the case of a URL also happens to tell us 
where the resource is.

I found the following slightly modified example from [t3n.de](https://t3n.de/news/url-uri-unterschiede-516483/) really helpful:

- website.com - URI
- https://website.com - URL
- ftp://website.com - URL

So URLs are not restricted to 'http' and 'https' as the example above shows. More generally, a URL consists of the following components:
$$
\underbrace{\text{http}}_{\text{protocol}}://
\underbrace{\text{www.website.com}}_{\text{host}}:
\underbrace{\text{42}}_{\text{port}}/
\underbrace{\text{some/path/to/resource}}_{\text{resource path}}
\underbrace{\text{?par1=val1&par2=val2}}_{\text{query}}
$$
If no port is set, the default for 'http' is 80 and 443 for 'https'.

The most common verbs for specifying http-requests are:

- **GET**: Get an existing resource
- **POST**: Create a new resource (usually with a payload that specifies the data used for the new resource)
- **PUT**: Update an existing resource
- **DELETE**: Delete a resource

When we submit an http-request, the server will respond with a status message:

- 1xx - informational 
- 2xx - success
- 3xx - redirection
- 4xx - client error (such as '404 - page not found', '401 - Unauthorized', or '403 - Forbidden')
- 5xx - server error

An http-message has the following defined structure:

1. Start line: Request line (POST) | Status line (GET)
2. One or more headers: Field name : Field value
    - Possible types: Request/response, general and entity headers
3. CRLF: Mandatory new line between headers and body
4. optional message-body

So an http-request could look like this:
> GET /some/path/to/resource HTTP/1.1
> Host: http:://www.website.com
> Connection: keep-alive
> Cache-Control: no-cache
> Pragma: no-cache
> Accept: text/html,application/xhtml+xml,application/xml;q=0.

As you can see, there are lots os headers, but only the 'HOST' header is mandatory. Since we submit a GET request, there 
is no message body. 

The response could look like this:

> HTTP/1.1 200 OK
> <response-headers>
>
> <body>

An excellent article wiht more detailed information about the http-protocol can be found [here](https://code.tutsplus.com/tutorials/http-the-protocol-every-web-developer-must-know-part-1--net-31177). If you are interested in 
reading about HTTP/2 check out this [article](https://developer.mozilla.org/en-US/docs/Web/HTTP/Messages) from Mozilla.

So, let's get started with some coding:)

## REST Calls in R with `httr`

The following section is based on Hadley Wickhams's excellent [httr-vignette](https://cran.r-project.org/web/packages/httr/vignettes/API-packages.html). Let's begin with simple API calls that do not need authentication. We will make the following REST call in R using httr:

```
GET /repos/hadley/httr HTTP/1.1
Host: api.github.com
Accept: application/vnd.github.v3+json
```

We call Github's public api and request information for Hadley's httr repo.

Using `httr` we can submit the request as follows:
```{r}
host = "https://api.github.com"
url = "/repos/hadley/httr"
resp = httr::GET(paste0(host, url))

resp
```

The response contains the header first followed by the body (application json in this case). We can automatically parse it to an R object using `jsonlite` or access status codes and headers using `httr` in-build functions like so:
```{r}
httr::status_code(resp)

head(httr::headers(resp), 3)

head(jsonlite::fromJSON(httr::content(resp, "text"), simplifyVector = FALSE), 3)
```

There is also an amazing website called [httpbin.org](httpbin.org) where you can try submitting requests:
```{r}
httr::POST(httr::modify_url("https://httpbin.org", path = "/post"))
```

Hadley recommends setting a user agent in your requests to identify the client that is calling the API so API owners can see who is using their services. By sending your R package as a user agent, API owners might also see that there is demand from R users and perhaps allocate more resources.

You can send a user agent like so:
```{r}
ua = httr::user_agent("http://github.com/hadley/httr")

resp = httr::GET("https://httpbin.org/get", ua)
resp
```

That's it for now. In my next post I want to cover OAuth 2.0, a very common authentication model used to secure API endpoints.

